---
title: "Data Wrangling with R"
subtitle: "IMB 320 Section 02"
author: "Austin Voth"
date: today
format: 
    html:
         theme: cosmo
         toc: true
         toc-depth: 3
         toc-location: right-body
         embed-resources: true
execute: 
  freeze: auto
---

> **Note**: This codebook is based of the content of the lecture notes published on [Data Wrangling in Tidyverse](https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r>=)

# Overview

## Learning Outcomes

By the end of this module, you will be able to:

-   Describe the concept of Data Wrangling.
-   Describe how Tibbles are different from data frames
-   Explain how to convert wide or long data to “Tidy” data
-   Explain how to merge relational data sets using join functions. (Next module)
-   Explain how to use grouped mutates and filter together.
-   Be familiar with major dplyr functions for transforming data.
-   Create a new variable with mutate() and case_when().
-   Use the pipe operator to shape the data to prepare for analysis and visualization

::: callout-tip
To use mouse multi-cursor editing feature do the following steps: 1. Select the text you want to edit 2. Press Shift + Alt + I to place a cursor at the end of each line. 3. You can now edit all the selected text simultaneously
:::

# Introduction to Data Wrangling

## Loding Packages

**Note**:

-   To add a code chunk, you can use the keyboard shortcut, but it depends on which mode you are in:

-   In visual mode: `Ctrl + Alt + I` (Windows/Linux) or `Cmd + Option + I` (Mac) in RStudio.

-   In source mode: `Ctrl + Shift + I` (Windows/Linux) or `Cmd + Shift + I` (Mac) in Rstudio.

```{r}
# install.packages("tidyverse")
# install.packages("nycflights13")
library(tidyverse)
library(nycflights13)
```

## Tibbles vs Data Frames

```{r}
class(mtcars)
cars.tib <- mtcars |>
  rownames_to_column() |>
  as_tibble() |>
  rename(model = rowname)
cars.tib
class(cars.tib)
typeof(cars.tib)
```

## Data Wrangling with dplyr

**Data sets in package ‘nycflights13’:**

-   airlines Airline names.
-   airports Airport metadata
-   flights Flights data
-   planes Plane metadata.
-   weather Hourly weather data

```{r}
data()
```

```{r}
flights_df <- flights

flights_df |>
  count(year, month) |>
  arrange(desc(n))  |>
  mutate(month = as_factor(month)) %>% 
  mutate(month = fct_reorder(month, n)) %>% 
  ggplot(aes(month, n, fill = year)) +
  geom_col(show.legend = FALSE) +
   coord_flip() +
  labs(x = "Month",
       y = "# of Flights",
       title = "# of flights by Month during year 2013 at New York") +
  theme_bw()
```

